Contracts for unions and intersections are hard for Nickel

### 2021-05-30 paper.tex
Pass on 1, 2

For instance, we could define a flat contract \nickel{even} as
\nickel{fun x => x \% 2 == 0}, that checks
whether a number is even by checking the remainder of dividing it by 2.
And then use it to check that a function always returns an
even number, when applied to an \nickel{Integer}, as shown in
Figure \ref{fig:num-to-even}.
Or even, we could come up with a contract that checks that a function
maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true},
and use it as a contract for the \nickel{not} function, as can be seen on Figure
\ref{fig:checked-not}.

\begin{figure}[h]
\begin{center}
\begin{lstlisting}[language=nickel]
let f = (fun n => n * 2) @ (Integer -> even)
\end{lstlisting}
  \end{center}
\caption{A function wrapped with a contract stating that it always returns
an even number}
\label{fig:num-to-even}
\end{figure}

% $$let~f~=~(\lambda n. ~ n * 2)@(Number ~ \rightarrow ~ even?)$$

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let notC = fun f => (f true == false) and (f false == true) in
let checkedNot = not @ notC
\end{lstlisting}
\end{center}
\caption{A contract that checks the not function behaves as expected}
\label{fig:checked-not}
\end{figure}

This type of contracts are present on many different languages,
for instance, the Eiffel programming language\unsure{Citation?}, the precursor
of the Design by Contract ideals, allows to assert
these kinds of expression as pre- and post-conditions on
functions\cite{EiffelDesignByContract}.
A somewhat newer language, the D programming
language, allows, among other things, to have invariants on
classes\unsure{I (Arnaud) believe Eiffel has this feature as well,
  doesn't it?},
written with flat contracts, that are
dinamically checked at the frontiers of every public method \cite{DLangContracts}.

Finally, the Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.
\todo{Talk about this, probably later on the paper.
See Related Works section}

@@

Beyond CSE, optimizations such as loop-invariant code motion or
let-floating~\ref{} apply the same principle of extracting out an invariant
expression to avoid its recomputation respectively across loop iterations and
function calls.

@@

Strikingly, we will see in section~\ref{} that the introduction of unions and
intersections can make this equivalence \todo{name it} unsound \emph{even in a
pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

\subsection*{Lazyness}
Evaluation strategies relates to the evaluation of the expression bound to
identifiers. Expressions are be bound by assignments (e.g. let bindings in
Nickel), or by function applications (arguments are bound to parameters). We
will use Figure~\ref{} as a working example.

@@

The eager strategy has been chosen in the vast majority of languages: while
theoretically less efficient, it is simpler to implement and more predictable
for the programmer. Also, lazy evaluation can occur performance overhead and
memory leaks in unexpected ways. While eager evaluation is the default, some
languages however offer a form of lazy evaluation, either as an opt-in option
(e.g. Ocaml and ReScript), or by default for specific primitives types such as
like list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
(and its web avatar PureScript). Although rare in general purpose languages,
lazy evaluation is more widespread in the configuration space: the configuration
languages Dhall, CUE and Jsonnet also implements full lazy evaluation. So are
the Nix language, which inspired Nickel, and Nickel itself.

@@

\section{Union \& intersection}
\label{sec:union-inter}
\info{What are they, what are they used for.}

>>>>

For instance, the contract \nickel{Even} shown on Figure
\ref{fig:contract-higher-order} is a flat contract, defined
by taking the value that needs to be checked, and validating
that is both a number and that the remainder of dividing it by 2
is 0.
And then, as seen on the same figure, we can use it to check that a function always
returns an \nickel{Int} when applied to an \nickel{Even}.

% \begin{figure}[h]
% \begin{center}
% \begin{lstlisting}[language=nickel]
% let f = (fun n => n * 2) @ (Integer -> even)
% \end{lstlisting}
%   \end{center}
% \caption{A function wrapped with a contract stating that it always returns
% an even number}
% \label{fig:num-to-even}
% \end{figure}

Or even more interesting, we could come up with a contract that checks that a function
maps \nickel{true} to \nickel{false} and \nickel{false} to \nickel{true}
by applying the function and checking its results,
and then use it as a contract for the \nickel{not} function, as can be seen on Figure
\ref{fig:checked-not}.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let NotC = contracts.fromPred (
  fun f => builtins.isFun f && (f true == false) && (f false == true)
) in
let checkedNot = not | #NotC
\end{lstlisting}
\end{center}
\caption{A contract that checks the not function behaves as expected}
\label{fig:checked-not}
\end{figure}

This type of contracts are present on many different languages,
for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
of the Design by Contract ideals, allows to assert
these kinds of expression as pre- and post-conditions on
functions and as invariants on classes\cite{EiffelDesignByContract}.
\resolved{I (Arnaud) believe Eiffel has this feature as well,
  doesn't it? (Teo) You're right},

% A somewhat newer language, the D programming
% language, allows, among other things, to have invariants on
% classes
% written with flat contracts, that are
% dinamically checked at the frontiers of every public method \cite{DLangContracts}.

Finally, the Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts
\footnote{And a gradual typing system
built on top of that, but that goes beyond the scope of this
work.}
, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.

@@

Beyond CSE, optimizations such as loop-invariant code motion or
let-floating~\cite{letFloating} apply the same principle of extracting out an invariant
\unsure{Was this the inteded citation?}
expression to avoid its recomputation respectively across loop iterations and
function calls.

@@

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of unions and
intersections can make this equivalence \todo{name it} unsound \emph{even in a
pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

\subsection*{Lazyness}
\todo{Lazyness does not appear on sec 4, if we don't put somethign there, I'd remove it}
Evaluation strategies relates to the evaluation of the expression bound to
identifiers. Expressions are bound by assignments (e.g. let bindings in
Nickel), or by function applications (arguments are bound to parameters). We
will use Figure~\ref{} as a working example.

@@

The eager strategy has been chosen in the vast majority of languages: while
theoretically less efficient, it is simpler to implement and more predictable
for the programmer. Also, lazy evaluation can occur performance overhead and
memory leaks in unexpected ways. While eager evaluation is the default, some
languages however offer a form of lazy evaluation, either as an opt-in option
(e.g. Ocaml and ReScript), or by default for specific primitives types such as
list (e.g.  Swift and Perl6). The most notable lazy language is Haskell
(and its web avatar PureScript). Although rare in general purpose languages,
lazy evaluation is more widespread in the configuration space: the configuration
languages Dhall, CUE and Jsonnet also implements full lazy evaluation. So are
the Nix language, which inspired Nickel, and Nickel itself.

@@

\section{Union \& intersection}
\label{sec:union-inter}
\info{What are they, what are they used for.}
\todo{Since nickel uses | for contract application, I'd move unions
to some other symbol.
I think the best option is to move to $\cap / \cup$}
